\documentclass{article}

% Core math + hyperlinks (does not change fonts/margins)
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{todonotes}
\usepackage{float}

% ---- Metadata (edit) ----
\title{CMDO 2025 - Sports Tournament Scheduling
(STS) problem}
\author{%
Luca Anzaldi \texttt{luca.anzaldi@studio.unibo.it} \and
Martino Michelotti \texttt{martino.michelotti@studio.unibo.it} \and
Alex-Cristian Cozma \texttt{alexcristian.cozma@studio.unibo.it}
}
\date{\today}

% ---- Helpful macros (optional) ----
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\opt}{\mathrm{opt}}
\newcommand{\AllDiff}{\texttt{allDifferent}}

\begin{document}
\maketitle

\section{Introduction}

\input{introduction}

% --------------------------------- SAT ------------------------------------------------------
\section{SAT Model}

\input{sat}


% --------------------------------- MIP ------------------------------------------------------
\section{MIP Model}

\input{mip}


% ---------------- CP ----------------
\clearpage
\section{CP Model}

\input{cp}


% --------------- Conclusions ---------------
\section{Conclusions}
The diversity of the programming languages and their limitations and strengths gave us different views on the problem and tools to handle it, letting us develop different models, having diverse variables and therefore also diverse constraints; without limiting ourselves to just develop three implementations of the same model. 

Among all models we developed, unless one specific case ($n$=18 with Constraint Programming), we had a good improvement for each of them when symmetry breaking (SB) constraints were applied.
For all three models, the respective SB constraints allowed us to find a solution when the problem was scaled further by increasing $n$ by one (+2) step.
Execution times were not reported, but they also had an improvement by adding SB constraints, outputting faster solutions by reducing the search tree.

We initially implemented constraints that fixed the home and away balances to directly force them on the optimal values. The mentioned constraints are
\[
\sum_{t=1}^{n}\big| \; 
\#\text{home\_games}(t) - \#\text{away\_games}(t) \; \big| = n.
\]
or alternatively
\[
\forall\, t \qquad \big| \; 
\#\text{home\_games}(t) - \#\text{away\_games}(t) \; \big| = 1
\]

By strictly constraining the objective function, the model was finding a solution (optimal by problem definition) in significant less time; but this would've changed the optimization problem in a satisfaction problem, so we kept them away to let the solver find the optimal solution without external hints.

The \textbf{SAT} approach has a clear limitation in its restriction to purely Boolean variables, which significantly constrained how we could define decision variables. Despite the considerable effort of implementing and testing four different cardinality encodings, the improvements in the final results remained limited for our formulation.
The SAT experiments highlighted that, in our case, performance is influenced more by the modeling choices than by symmetry-breaking techniques. While symmetry breaking yielded incremental improvements, reformulating the model produced substantially larger performance gains.

\textbf{Linear Programming} allowed us to set integer variables in the problem, but it would have been trickier to define the constraints, especially when summing up variables. Further issues encountered with MIP were the linear limitations we had defining functions and constraint, bringing us to define new auxiliary variables and reformulate them as linear inequalities.

\textbf{Constraint programming} was for sure the more flexible method, admitting in the formulation every function we needed and different kind of variables; but it's exactly here where we had to think and experiment plenty of different options to find the best combination that could facilitate the constraint propagation of the solver and consequently the reduction of the search tree.

% --------------- Authorship ---------------
\newpage
\section*{Authenticity and Author Contribution Statement}
We declare that the work reported here is our own and properly cites all external ideas and sources.
Parts of this project (text/code/experiments) were assisted by AI tools where indicated; we disclose the tools used and the exact sections where AI assistance was applied expect for the part where it was used to improve the quality of the report, such as grammar, sentence formulation and for code debugging.
\\[0.5em]

\paragraph{Author contributions.}
Luca Anzaldi developed the SAT-based formulations, including both Model~1 and
Model~2. He was responsible the
configuration and setup of all scripts, Docker environments, and auxiliary
tools required to run the experiments.
Martino Michelotti developed the complete CP model.
The report was written by both students, mostly reporting the part of the project they were responsible of.

% --------------- References ---------------
\begin{thebibliography}{9}

\bibitem{minizinc}
MiniZinc: The Modeling Language.\\
\url{https://www.minizinc.org/}

\bibitem{csplib_prob026}
CSPLib, \emph{Problem 026: Sports Tournament Scheduling (prob026)}.\\
\url{https://www.csplib.org/Problems/prob026/}
(Accessed: 2026-01-29)

\bibitem{z3}
Leonardo de Moura and Nikolaj Bj{\o}rner.
\emph{Z3: An Efficient SMT Solver}.
In \emph{Proceedings of the 14th International Conference on Tools and Algorithms for the Construction and Analysis of Systems (TACAS)}, 
LNCS 4963, pp. 337--340, Springer, 2008.\\
\url{https://github.com/Z3Prover/z3}

\bibitem{pythonmip}
Henrique R. Lopes, Thiago A. F. S. Santos, and Tallys H. Y. Lopes.
\emph{Python-MIP: A Python Toolbox for Modeling and Solving Mixed-Integer Linear Programs}.
SoftwareX, Volume 12, 2020.\\
\url{https://www.python-mip.com/}


\end{thebibliography}

\end{document}
